The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-02-14 dictionary.tex
Changes for submitting to Algebra & Number Theory

% !TEX encoding = UTF-8 Unicode
%\documentclass{compositio}
\documentclass{amsart}
\pdfoutput=1

%%%%%%%%%%%%%%%%%% HEADING %%%%%%%%%%%%%%%%%%
\title[The function--sheaf dictionary]{The function--sheaf dictionary for smooth commutative group schemes over finite fields}
% with finitely generated geometric component group}
\date{\today}
\author{Clifton Cunningham}
\email{cunning@math.ucalgary.ca}
\address{University of Calgary}
\author{David Roe}
\email{roed.math@gmail.com}
\address{Pacific Institute for the Mathematical Sciences at the University of Calgary}
%\classification{14L15 (primary), 11G25, 14F05, 14D24, 11S31 (secondary)}
%\keywords{quasicharacters, quasicharacter sheaves, algebraic tori, Greenberg transforms, N\'eron models, local fields, group schemes}

@@

The sheaf--function dictionary for connected commutative algebraic groups over finite fields \cite{deligne:SGA4.5}*{sommes trig.} translates characters into local systems. Let $G$ be a commutative group scheme over a finite field $\Fq$, let $\ell$ be any prime invertible in $\Fq$ and let $\chi : G(\Fq) \to \EEx$ be any character. Using little more than the Lang map for $G$, $\chi$ determines an $\ell$-adic local system $\qcs{L}$ on $G$ from which the character $\chi$ can be recovered using the Frobenius automorphism of $\bG$, the base change of $G$ to $\bFq$. While this recovery process applies to all $\ell$-adic local systems $\qcs{L}$ on $G$, those that produce characters of $G(\Fq)$ are distinguished by a simple property: there exists an isomorphism 
\begin{equation}\label{introbox}
m^* \qcs{L} \iso \qcs{L} \boxtimes \qcs{L}
\end{equation}
 of $\ell$-adic local systems on $G$, where $m : G\times G \to G$ is the multiplication map for $G$. The sheaf--function dictionary for $G$ identifies a monoidal category of $\ell$-adic local systems on $G$ such that the group of isomorphism classes in this category is canonically identified with the $\ell$-adic character group of $G(\Fq)$, thus providing the geometrization and categorification of characters of $G(\Fq)$.

Although hidden in the discussion above, the connectness of $G$ played a crucial role in the geometrization and categorification of characters of $G(\Fq)$.  So it is natural to ask if the sheaf--function dictionary can be generalized to not-necessarily connected commutative algebraic groups over finite fields. In this paper we answer that question. 
%
In fact, we show much more by exhibiting a sheaf--function dictionary, in the sense above, for any smooth commutative group scheme $G$, locally of finite type over a finite field $\Fq$, 
for which the geometric component group of $G$ is finitely generated; we call such group schemes $\QC$-schemes.
To do this we must pass from $\ell$-adic local systems on a $\QC$-scheme $G$ to Weil local systems on $G$ 
and also keep track of the isomorphisms \eqref{introbox} (existence of such an isomorphism is not enough), leading to the notion of {\it quasicharacter sheaves on $G$}.

@@

A simple example will illustrate some of the issues that arise when extending the sheaf--function dictionary to $\QC$-schemes. 
Consider the etale group scheme $G = \ZZ_{/\Fq}$. 
Characters of $G(\Fq) = \ZZ$ are neither finite nor bounded, in general. 
For example, while the $\ell$-adic character $\ZZ \to \EEx$ defined by 
$1 \mapsto -1$ has finite order, the character defined by 
$1 \mapsto 1+ \ell$ does not, although its image is bounded, while the character $\ZZ \to \EEx$ defined by 
$1 \mapsto \ell$ has unbounded image.  
As we show in the paper, $\ell$-adic local systems on $G$ can only be used to geometrize characters of finite order; quasicharacter sheaves, however, provide the geometrization of all characters of $G(\Fq)$.

>>>>

% !TEX encoding = UTF-8 Unicode
\documentclass{amsart}
\pdfoutput=1

%%%%%%%%%%%%%%%%%% HEADING %%%%%%%%%%%%%%%%%%
\title[The function--sheaf dictionary]{The function--sheaf dictionary for smooth commutative group schemes over finite fields}
% with finitely generated geometric component group}
\date{\today}
\author{Clifton Cunningham}
\email{cunning@math.ucalgary.ca}
\address{University of Calgary}
\author{David Roe}
\email{roed.math@gmail.com}
\address{Pacific Institute for the Mathematical Sciences at the University of Calgary}
\subjclass[2010]{14F05 (primary), 14L15 (secondary)}
\keywords{sheaf--function dictionary, geometrization, quasicharacter sheaves, commutative group schemes}

@@

The function--sheaf dictionary for connected commutative algebraic groups over finite fields \cite{deligne:SGA4.5}*{sommes trig.} translates characters into local systems. Let $G$ be a commutative group scheme over a finite field $\Fq$, let $\ell$ be any prime invertible in $\Fq$ and let $\chi : G(\Fq) \to \EEx$ be any character. Using little more than the Lang map for $G$, $\chi$ determines an $\ell$-adic local system $\qcs{L}$ on $G$ from which the character $\chi$ can be recovered using the Frobenius automorphism of $\bG$, the base change of $G$ to $\bFq$. While this recovery process applies to all $\ell$-adic local systems $\qcs{L}$ on $G$, those that produce characters of $G(\Fq)$ are distinguished by a simple property: there exists an isomorphism 
\begin{equation}\label{introbox}
m^* \qcs{L} \iso \qcs{L} \boxtimes \qcs{L}
\end{equation}
 of $\ell$-adic local systems on $G$, where $m : G\times G \to G$ is the multiplication map for $G$. The function--sheaf dictionary for $G$ identifies a monoidal category of $\ell$-adic local systems on $G$ such that the group of isomorphism classes in this category is canonically identified with the $\ell$-adic character group of $G(\Fq)$, thus providing the geometrization and categorification of characters of $G(\Fq)$.

Although hidden in the discussion above, the connectness of $G$ played a crucial role in the geometrization and categorification of characters of $G(\Fq)$.  So it is natural to ask if the function--sheaf dictionary can be generalized to not-necessarily connected commutative algebraic groups over finite fields. In this paper we answer that question. 
%
In fact, we show much more by exhibiting a function--sheaf dictionary, in the sense above, for any smooth commutative group scheme $G$, locally of finite type over a finite field $\Fq$, 
for which the geometric component group of $G$ is finitely generated; we call such group schemes $\QC$-schemes.
To do this we must pass from $\ell$-adic local systems on a $\QC$-scheme $G$ to Weil local systems on $G$ 
and also keep track of the isomorphisms \eqref{introbox} (existence of such an isomorphism is not enough), leading to the notion of {\it quasicharacter sheaves on $G$}.

@@

A simple example will illustrate some of the issues that arise when extending the function--sheaf dictionary to $\QC$-schemes. 
Consider the etale group scheme $G = \ZZ_{/\Fq}$. 
Characters of $G(\Fq) = \ZZ$ are neither finite nor bounded, in general. 
For example, while the $\ell$-adic character $\ZZ \to \EEx$ defined by 
$1 \mapsto -1$ has finite order, the character defined by 
$1 \mapsto 1+ \ell$ does not, although its image is bounded, while the character $\ZZ \to \EEx$ defined by 
$1 \mapsto \ell$ has unbounded image.  
As we show in the paper, $\ell$-adic local systems on $G$ can only be used to geometrize characters of finite order; quasicharacter sheaves, however, provide the geometrization of all characters of $G(\Fq)$.
